---
title: "How does a bike-share navigate speedy success?"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = getwd())
```

## An introduction to the case study

- I am a **junior data analyst** working in the **marketing analytics team** at Cyclistic. 
- **Cyclistic** is a fictional bike-share company with more than 5,800 bicycles and 600 docking stations across Chicago. 
- Cyclistic's bike-share program offers flexibility in pricing plans: *single-ride passes*, *full-day passes*, and *annual memberships*. 
- Customers who purchase single-ride passes or full-day passes are referred to as **casual riders**. 
- Customers who purchase annual memberships are **annual members**. 

## A clear summary of the business task

The primary objective for our team is to design effective marketing strategies that convert casual riders into annual members. To achieve this, we must address the following key questions: 

- How do annual members and casual riders use Cyclistic bikes differently?
- Why would casual riders buy Cyclistic annual memberships?
- How can Cyclistic use digital media to influence casual riders to become members?

My focus is on the first question, as understanding usage differences is foundational for informing subsequent strategies.

**Business Task:** Analyze Cyclistic's historical bike trip data to uncover trends and key differences in usage between annual members and casual riders. These insights will guide recommendations for targeted marketing initiatives aimed at increasing annual memberships.

## A description of all data sources used

- As Cyclistic is a fictional company, I am using “Divvy bicycle sharing service” [trip data](https://divvy-tripdata.s3.amazonaws.com/index.html) for this analysis. 
- “Lyft Bikes and Scooters”, which manages Chicago's Divvy bike-sharing program, shared specific Divvy system data to the public under this [license agreement](https://divvybikes.com/data-license-agreement). 
- The datasets are *reliable*, *original*, *comprehensive*, *current*, and *cited* thus proving data's **credibility** and **integrity**. 
- I will protect riders' privacy by actively preventing usage of any personally identifiable information. 
- I downloaded and stored data into GitHub for extra layer of security. 

## Documentation of any cleaning or manipulation of data

- Load tidyverse and conflicted packages to help wrangle data and manage conflicts.

```{r load-packages, echo=TRUE}
library(tidyverse) # helps wrangle data
library(conflicted) # to manage conflicts
# set dplyr::filter and dplyr::lag as the default choices
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

- Import “Divvy 2019 Q1” and “Divvy 2020 Q1” datasets. 

```{r import-datasets, echo=TRUE}
q1_2019 <- read_csv(
  "divvy-tripdata/Divvy_Trips_2019_Q1/Divvy_Trips_2019_Q1.csv",
  show_col_types = FALSE
)
q1_2020 <- read_csv(
  "divvy-tripdata/Divvy_Trips_2020_Q1/Divvy_Trips_2020_Q1.csv",
  show_col_types = FALSE
)
```

- Inspect column names in both data frames to see if they are consistent.

```{r inspect-column-names, echo=TRUE}
cat(
  "Column names of 2019 Q1 dataframe:",
  paste(colnames(q1_2019), collapse = ", "), "\n"
)
cat(
  "Column names of 2020 Q1 dataframe:",
  paste(colnames(q1_2020), collapse = ", "),
  "\n"
)
```

- Standardize column names of the data frames to make them consistent.

```{r standardize-columns, echo=TRUE}
q1_2019 <- q1_2019 %>%
  rename(
    ride_id = trip_id,
    rideable_type = bikeid,
    started_at = start_time,
    ended_at = end_time,
    start_station_name = from_station_name,
    start_station_id = from_station_id,
    end_station_name = to_station_name,
    end_station_id = to_station_id,
    member_casual = usertype
  )
```

- Inspect structure of both data frames to see if they are consistent.

```{r inspect-structures, echo=TRUE}
str(q1_2019)
str(q1_2020)
```

- Standardized data types of ride_id and rideable_type columns in both data frames to ensure consistency and avoid potential issues during data merging.

```{r standardize-types, echo=TRUE}
q1_2019 <- q1_2019 %>%
  mutate(
    ride_id = as.character(ride_id),
    rideable_type = as.character(rideable_type)
  )
```

- Combine both datasets into a single dataframe for analysis. 

```{r combine-dataframes, echo=TRUE}
all_trips <- bind_rows(q1_2019, q1_2020)
```

- Remove the fields that are not relevant for the analysis. The fields `tripduration`, `gender`, `birthyear`, `start_lat`, `start_lng`, `end_lat`, and `end_lng` are not needed for our analysis, so we will remove them.

```{r remove-irrelevant-fields, echo=TRUE}
all_trips <- all_trips %>%
  select(
    -c(tripduration, gender, birthyear, start_lat, start_lng, end_lat, end_lng)
  )
```

- List out the column names of the new data frame.

```{r df-column-names}
paste(colnames(all_trips), collapse = ", ")
```

- How many rows are in the data frame?

```{r df-rows-count}
nrow(all_trips)
```

- Show dimensions of the data frame.

```{r df-dimensions}
dim(all_trips)
```

- See the first 6 rows of the data frame.

```{r df-head}
head(all_trips)
```

- See the last 6 rows of the data frame.

```{r df-tail}
tail(all_trips)
```

- See the structure of the data frame.

```{r df-structure}
str(all_trips)
```

- Statistical summary of the data (mainly numeric).

```{r df-summary}
summary(all_trips)
```

- How many observations fall under each rider type?

```{r rider-types}
table(all_trips$member_casual)
```

- Harmonize rider type labels: replace "Subscriber" with "member" and "Customer" with "casual".

```{r harmonize-rider-types}
all_trips <- all_trips %>%
  mutate(member_casual = recode(member_casual,
                                "Subscriber" = "member",
                                "Customer" = "casual"))
table(all_trips$member_casual)
```

- Add new columns for ride duration, date, month, day, and year.

```{r add-new-columns}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(all_trips$date, "%m")
all_trips$day <- format(all_trips$date, "%d")
all_trips$year <- format(all_trips$date, "%Y")
all_trips$day_of_week <- wday(all_trips$started_at, label = TRUE)
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
```

- Inspect the structure of the columns.

```{r new-col-structure}
str(all_trips)
```

- Convert ride_length into numeric for data calculations.

```{r transform-ride-length}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

- Remove "bad" data. The data frame includes entries when bikes were taken out of docks and checked for quality by Divvy (with station name as "HQ QR") or ride_length was negative.

```{r remove-bad-data}
all_trips_v2 <- all_trips[
  !(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0),
]
cat(nrow(all_trips) - nrow(all_trips_v2), "rows with bad data are removed.\n")
```

## A summary of my analysis

- Calculated key metrics (average, median, maximum, minimum) for ride counts and average duration. 

```{r ride-length-summary}
summary(all_trips_v2$ride_length)
```

- Compare ride duration patterns by rider type (annual members vs. casual riders). 

```{r compare-ride-length}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

- Compare average ride duration by rider type (annual members vs. casual riders) and day of the week. 

```{r avg-ride-length-patterns}
aggregate(
  all_trips_v2$ride_length ~
    all_trips_v2$member_casual + all_trips_v2$day_of_week,
  FUN = mean
)
```

- Compare ridership (ride counts and average ride duration) patterns by rider type (annual members vs. casual riders) and day of the week. 

```{r ridership-patterns}
all_trips_v2 %>%
  group_by(member_casual, day_of_week) %>%
  summarise(
    number_of_rides = n(),
    average_ride_duration = mean(ride_length),
    .groups = "drop" # drops all levels of grouping
  ) %>%
  arrange(member_casual, day_of_week)
```

- Create column chart to visualize ride count patterns by rider type (annual members vs. casual riders) and day of the week. 

```{r ride-counts-viz}
all_trips_v2 %>%
  group_by(member_casual, day_of_week) %>%
  summarise(
    number_of_rides = n(),
    .groups = "drop"
  ) %>%
  arrange(member_casual, day_of_week) %>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of rides for Annual members VS Casual riders",
    x = "Day of the week",
    y = "Number of rides"
  )
```

- Create column chart to visualize average ride duration patterns by rider type (annual members vs. casual riders) and day of the week.

```{r average-ride-duration-viz}
all_trips_v2 %>%
  group_by(member_casual, day_of_week) %>%
  summarise(
    average_ride_duration = mean(ride_length),
    .groups = "drop"
  ) %>%
  arrange(member_casual, day_of_week) %>%
  ggplot(
    aes(x = day_of_week, y = average_ride_duration, fill = member_casual)
  ) +
  geom_col(position = "dodge") +
  labs(
    title = "Average ride durations for Annual members VS Casual riders",
    x = "Day of the week",
    y = "Average ride durations (in seconds)"
  )
```

- Export summary files for further analysis

```{r export-summary}
analysis_summary <- all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(
    number_of_rides = n(),
    average_ride_duration = mean(ride_length),
    .groups = "drop"
  ) %>%
  arrange(member_casual, weekday)
write.csv(analysis_summary, file = "analysis-results/analysis_summary.csv")
```

## Supporting visualizations and key findings

- Annual members consistently take more rides than casual riders throughout the week.
- On weekdays, annual members take significantly more rides compared to weekends.
- The average ride duration for annual members remains similar across all days.
- Casual riders have longer average ride durations than annual members on any day.
- Casual riders take more rides on weekends than on weekdays.

- Check out this interactive [Tableau dashboard](https://public.tableau.com/views/CyclisticBike-shareAnalysis_17481008517970/CyclisticBike-shareAnalysis) to visualize the analysis results using the exported summary file.

## My top three recommendations based on my analysis

1. **Annual members ride more frequently:** Annual members consistently take more rides than casual riders throughout the week, especially on weekdays.
2. **Casual riders prefer weekends and longer rides:** Casual riders take more rides on weekends and have longer average ride durations than annual members on any day.
3. **Ride duration is stable for annual members:** The average ride duration for annual members remains similar across all days, indicating consistent usage patterns.

## Possible next steps for the case study

- Currently, 8% of ridership comes from users who depend on assistive options. However, the existing proxy data lacks sufficient detail for a comprehensive analysis of this group. I can seek additional resources to address this gap. 
- Additionally, I can collaborate with the team to create a targeted marketing strategy focused on converting casual riders into annual members.